---
title: "HDDA_Bonus_Ex4"
author: "Joshua Damm"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
load("exercise2_grp28.RData")
```

1. You are given a list of p-values (object named pall). What is the 3rd smallest Benjamini-Hochberg corrected p-value?

```{r}
# sort pvals
pval_sorted = sort(pall)
# total number of tests
m = length(pval_sorted)

# desired false discovery rate
FDR = q = 0.05

# Initialize vectors
crit_val <- numeric(m)
BH_adjust <- numeric(m)
decision <- logical(m)

# Loop to calculate thresholds
for (i in 1:m) {
  crit_val[i] <- i / m * q
  }
  

for (i in 1:m) {
    
  # Make decision based on comparison
    if (pval_sorted[i] <= crit_val[i]) {
      decision[i] <- TRUE
    }
    
    else {
      decision[i] <- F
    }
    
    # adjust p-values
    BH_adjust[i] = pmin((m/i) * pval_sorted[i] , 1)
    
}

answer1 = BH_adjust[3]
answer1

```

```{r}
p_adjusted = p.adjust(pall, method = "BH", n = length(pall))

sort(p_adjusted)

```


From the same list of p-values, estimate the lfdr at cutoff t=0.03. Assume the null-distribution is Uniform[0,1]. Use the density function in R to estimate the density of p-values. Use defaults for the arguments, but take into account that p-values are bounded. Estimate pi_0 by assuming f_1(1) = 0 and using the mixture model f(p) = pi0 f_0(p) + (1-pi0)f_1(p)




```{r}
# Estimate the density of all raw p-values
p_density <- density(pall, from = 0, to = 1)

# plot mixture distribution
plot(p_density$x, p_density$y)

# Calculate f(1) and estimate pi_0
f_1 <- approx(p_density$x, p_density$y, xout = 1)$y  # Density at p = 1
pi_0 <- f_1  # Estimate of pi_0

# Calculate f(t) at t = 0.03
f_t <- approx(p_density$x, p_density$y, xout = 0.03)$y  # Density at t = 0.03

# Step 4: Calculate lFDR at t = 0.03, since f0(0.03) = 1
lfdr_t <- pi_0 / f_t

# Display result
answer2 = lfdr_t

answer2
```


# Question 3

You are given a (nnull x p) matrix of p-values (object named pmat), where each column represents a test item and each row a realization of the p-value vector under the joint null-hypothesis (null-distribution may deviate from Uniform[0,1]). From the list of pvalues and this matrix, estimate the FDP for threshold t = 0.01, assuming pi0 = 1

use the entire matrix to estimate F_0(t=0.01)

```{r}
# get pooled p-value vector by pooling permutations, since i.i.d
p_pooled = c(pmat)



t = 0.01

m = ncol(pmat)

pi_0 = 1


# get cdfs

# cdf pooled, H0
F_0 = ecdf(p_pooled)
#Plot the CDF
plot(F_0, 
     main = "Empirical CDF",
     xlab = "pvals",
     ylab = "Cumulative Probability",
     col = "blue",
     lwd = 2) 



FDP_est = pi_0 * m * F_0(t) / (sum(p_pooled <= t))

answer3 = FDP_est

answer3
```




